---
title: "Untitled"
output: html_document
---

# Part 1: Data Preparation

The missing rate of  HDI.for.year in master.CSV is 69.93%, so delete this variable and select the data from 1996 to 2016 for analysis. In the Economic data, four indicators, namely "Services, value added (% of GDP)", "Industry (including construction), value added (% of GDP)", "Manufacturing, value added (% of GDP)", "Agriculture, forestry, and fishing, value added (% of GDP), were selected to analyze their impact on suicide rate.

```{r setup, include=FALSE}
#my file location: C:\Users\a5730\Downloads\380_final
setwd("C:\\users\\a5730\\Downloads\\380_final")

#Suicide rate Dataset
master <- read.csv("master.csv", 
                   header = T, 
                   sep = ",",
                   stringsAsFactors = F)
#Economic dataset
Economic <- read.csv("API_3_DS2_en_csv_v2_10523844.csv",
                     header = T, 
                     sep = ",", 
                     stringsAsFactors = F)
```

```{r}
# missing rate
colSums(is.na(master))/nrow(master)
# delete HDI.for.year and country.year
master <- master[which(master$year>1996 & master$year<2017),-c(8,9)]
```

```{r}
# Economic data processing
fun1 <- function(x){
  data <- Economic[which(Economic$Indicator.Code==x),c(1,4,41:61)]
  data2 <- data[,c(1,3)]
  data2$year <- 1996
  colnames(data2) <- c("country",x,"year")
  j=1
  for(i in 4:23){
    data1 <- data[,c(1,i)]
    data1$year <- 1996+j
    colnames(data1) <- c("country",x,"year")
    data2 <- rbind(data2,data1)
    j=j+1
  }
  return(data2)
}

value <- c("NV.SRV.TOTL.ZS","NV.IND.TOTL.ZS","NV.IND.MANF.ZS","NV.AGR.TOTL.ZS")
Economic_Services <- fun1(value[1])
Economic_Industry <- fun1(value[2])
Economic_Manufacturing <- fun1(value[3])
Economic_Agriculture <- fun1(value[4])

Economic2 <- merge(Economic_Services, Economic_Industry, 
                   all = T, 
                   by=c("country","year"))
Economic2 <- merge(Economic2, Economic_Manufacturing,
                   all = T, 
                   by=c("country","year"))
Economic2 <- merge(Economic2, Economic_Agriculture,
                   all = T,
                   by=c("country","year"))
```


```{r}
# merge master and Economic2
master <- merge(master, Economic2,
                all.x = T,
                by=c("country","year"))
# missing rate
colSums(is.na(master))/nrow(master)
# delete missing data
master <- na.omit(master)
```

# Part 2: Data Introduction

